---
title: "Capstone Project Report"
subtitle: "Movie Similarity"
author: "David Pinto"
date: "`r Sys.Date()`"
header-includes:
  \renewcommand{\contentsname}{Contents}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
output: 
   hrbrthemes::ipsum_pdf:
      toc: true
      toc_depth: 2
      number_sections: false
      highlight: tango
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,  dev="cairo_pdf", message=FALSE, warning=FALSE,
                      fig.align="center", fig.path="img/", dpi = 90, fig.height = 4.25, 
                      out.width = "80%", cache = TRUE, cache.path = "cache/")
```

# I. Definition

## Project Overview

The main goal of this project is to develop a **Movie Similarity Recommender** over the full version of the famous [MovieLens](https://grouplens.org/datasets/movielens/) dataset. The developed solution aims to recommend similar items given a particular movie, so that users that already watched the later are likely to enjoy the recommended movies. It involves **Recommender Systems**$^1$ theory and **Similarity Matching**$^2$ techniques.

To properly achieve this we combine two datasets, one that provides movie ratings to be used as interaction data, and another that provides movie metadata to be used as content features.

### Movie Ratings

The full [MovieLens](https://grouplens.org/datasets/movielens/latest/) dataset, which contains `27M ratings` given to `58K movies` by `280K users`.

### Movie Metadata

Data retrieved from the [TMDB Open API](https://www.themoviedb.org/documentation/api) and available as a [Kaggle Dataset](https://www.kaggle.com/rounakbanik/the-movies-dataset). It contains information on 45k movies featured in the full MovieLens dataset. Features include title, description, genre and cast&crew information.

## Problem Statement

Recommending similar items is a very common task in many online selling platforms$^3$. It provides recommendations like: **related to items you've viewed**, **recommended for you based on your previous purchase**, **customers who bought this item already bought**, so on. The main challenges involved are:

**(a) The cold start problem**

How to recommend recent added products with few or no associated interaction data?

One way to cope with this consists in take advantage of product metadata (content features): `name`, `description`, `category`, `price level`, so on. Then we can link a product with others by content similarity. We call this as **Content Based** approach.

**(b) The popularity bias problem**

According to the **Long Tail phenomenon**$^4$, selling niche products has many advantages. Users are likely to pay more for niche products because they cannot find it everywhere. However, best seller products have more interaction data available to train a recommender system so that the fitted model may be subjected to popularity bias. In other words, the model will be more likely to recommend popular items.

Again, using content features to find similar items reduces popularity bias. But interaction data is very important to understand user behavior and cannot be discarded. So, in this project we apply a **Collaborative Filtering** factorization that takes into account item and user popularity.

### Movie Embeddings and Similarity Matching

First of all, we use both interaction and content data to extract item embedding$^5$ features, which are low-dimensional dense vectors that captures some of the semantics of the item by placing semantically similar items close together in the embedding space. It allows efficient nearest neighbors search, so that we can find similar movies fast in application time.

- **Content Embedding:** embedding features extracted from text fields like title and description, and from categorical fields like genre and cast&crew.
- **Collaborative Filtering Embedding:** embedding features extracted from user-movie interactions given by the ratings.

All the details about the feature extraction process will be discussed later.

After computing the embedding features, similar movies will be recommended using **Approximate Nearest Neighbors** (ANN) search with cosine similarity. The cosine similarity is the most common affinity metric applied for similarity search in the embedding space.

## Metrics

### Users Intersection as a Similarity Measure

Let's consider every rating (from 1 to 5) above 3.5 as an explicit feedback that the user liked the movie. Then we can measure the similarity between a pair of movies $(x,y)$ as:

$$
S(x,y) = \displaystyle\frac{N_{xy}}{\sqrt{N_xN_y}}
$$

where $N_x$ is the number of users that liked movie x, $N_y$ is the number of users that liked movie y, and $N_{xy}$ is the number of users that liked both x and y.

So, $S(x,y)$ close to 1 is a strong evidence that users who liked x will like y too. We call this as **User Correlation**.

Despite providing good recommendations, this similarity is very sparse, resulting in a discrete recommendation. So, if we try to recommend the top 30 similar to each movie we will get less than 30 for many movies. In other words, we will have many movies with few or none similar items to be recommended. 

The embedding features solve this problem. Since we are computing distances between pairs of movies in a dense and low-dimensional space, the embedding features provide a continuous recommendation. We are able to find similar items to any movie, with high evidence (low distance) or low evidence (high distance).

### Evaluating the Embedding Features

How close the similarity match with embedding features can get to the similarity match given by the user's likes (user correlation)?

**(a) Ranking correlation**

In a attempt to answer this question, we can randomly select $n$ test movies and get their top N similar using the $S(x,y)$ measure. Then, for each test movie, we get the distances $D(x,y)$ from the top N similar using the embedding features. Finally, for each test movie, we apply the Spearman's Rank-Order Correlation Coefficient ($\rho$) between $S$ and $1/D$, which is equal to the Pearson's correlation between the rank values of them.

For all test movies, we get an average rank correlation:

$$
\frac{1}{n}\sum_{i=1}^n\rho(S_i, 1/D_i)
$$

**(b) Precision and Recall at K**

Given a threshold $t_S$ for $S(x,y)$, we will assume that $S(x,y) \geq t_S$ indicates a strong/relevant similarity. So, given the K nearest neighbors from embedding features of a test movie:

$$
\text{precision@k} = \frac{\#\text{relevants@k}}{K}
$$

$$
\text{recall@k} = \frac{\#\text{relevants@k}}{\#\text{relevants}}
$$

For all test movies, we get an average $\text{Precision@K}$ and an average $\text{Recall@K}$.

# II. Analysis

## Data Preparation

First of all, take a look at `data/raw` to get instructions on how to download the datasets.

Data Preparation was a quite hard extra work in this project. As a previous step, we had to apply some manipulations in order to prepare data for analysis: 

1. We set MovieLens ID as the primary key in both datasets, MovieLens and TMDB.
1. We transformed Cast&Crew, movie Genre and movie Keywords data from a JSON format to a [Tidy Data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) format.
1. We concatenated all movie Keywords in a single text field.
1. We concatenated all movie Tags assigned by the users in a single text field.

Then, we applied some filters in an attempt to discard weak relations in the data:

1. In the Cast data we selected actors that appeared in at least 10 movies.
1. In the Crew data we considered just `director`, `producer` and `writer`, and selected those that appeared in at least 10 movies.
1. From the movie metadata we selected just the text fields: `title`, `overview` and `tagline`.
1. From the ratings data (interaction data) we removed users which voted less than 3 tines and movies with less than 30 votes. Items and users with poor interaction information can induce Collaborative Filtering to learn wrong correlations in the data.

All code is well organized and documented in the `01-data-preparation.ipynb` notebook.

## Data Exploration

From the Data Preparation step we finished up with 7 datasets in tidy format:

- `data/movie_info.csv`:
```{r}
library(data.table)
library(kableExtra)

dt <- fread("../data/movie_info.csv", nrows=5, select=c("id", "title", "overview", "tagline"))
dt[, overview := sprintf("%s (...)", substr(overview, 1, 30))]
dt[, tagline := sprintf("%s (...)", substr(tagline, 1, 30))]

kbl(dt, booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

- `data/movie_keywords.csv`:
```{r}
dt <- fread("../data/movie_keywords.csv", nrows=5)
dt[, keywords := sprintf("%s (...)", substr(keywords, 1, 30))]

kbl(dt, booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

- `data/movie_tags.csv`:
```{r}
dt <- fread("../data/movie_tags.csv", nrows=5)
dt[, tags := sprintf("%s (...)", substr(tags, 1, 30))]

kbl(dt, booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

- `data/movie_genre.csv`:
```{r}
dt <- fread("../data/movie_genre.csv", nrows=5)

kbl(dt, booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

- `data/movie_actor.csv`:
```{r}
dt <- fread("../data/movie_actor.csv", nrows=5)

kbl(dt, booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

- `data/movie_producer`:
```{r}
dt <- fread("../data/movie_producer.csv", nrows=5)

kbl(dt, booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

- `data/movie_ratings`:
```{r}
dt <- fread("../data/movie_ratings.csv", nrows=5)

kbl(dt, booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

After all cleaning process we got text description for `20400` movies, keywords for `31116` movies, tags for `45981` movies, genres for `20136` movies, `10465` actors for a total of `35649` movies, `4106` director/producer/writer for a total of `24325` movies, and `27516829` ratings for a total of `15967` movies given by `273643` distinct users.

## Exploratory Visualization

Now we invite you to take a brief tour over the data. For more details, the `02-exploratory-analysis.ipynb` notebook contains all code for the Exploratory Data Analysis we present here.

### Ratings Distribution

First of all, let's observe the distribution of ratings. We have 10 possible values for the user rating: `[0.5, 1.0, 1.5, 2.0, 2.5, 3.0, 3.5, 4.0, 4.5, 5.0]`. In which proportion each one of these values appear in the data?

```{r}
library(symplyr)
library(data.table)
library(ggplot2)

ratings <- file_import_csv("../data/movie_ratings.csv", threads = 6)

ratings[, rating_cat := as.factor(sprintf('%.01f', rating))]
ratings.dist <- ratings[, .(n = .N), by = 'rating_cat']

plot_bars(x = ratings.dist$n, labels = ratings.dist$rating_cat) +
  labs(x = "Rating Level", y = "Frequency") +
  theme_symplyr()
```

Just a small part of the ratings fall bellow `3.0`. If we look at the movie average rating we will observe a similar behavior:

```{r}
ratings.avg = ratings[, .(avg_rating = mean(rating)), by = 'id']

plot_histogram(x = ratings.avg$avg_rating, breaks = 30) +
  labs(x = "Average Rating", y = "Frequency") +
  scale_x_continuous(breaks = seq(0, 5, by = 0.5)) +
  scale_y_continuous(breaks = seq(0, 1.5e3, by = 100)) +
  theme_symplyr()
```

In general, users tend to give good ratings to the movies they vote. It may be an evidence that users are most likely to vote in those movies they like. It makes sense, since users usually watch movies with actors and/or directors they already know and like, and in the genre they like most.

### Ratings as Likes and Dislikes

In this project we will convert ratings to likes (`>=3.5`) and dislikes (`<3.5`). Let's take a look the the distribution of number of likes:

```{r}
ratings[, like := ifelse(rating >= 3.5, 1, 0)]
ratings.like = ratings[, .(avg_rating = mean(rating), num_likes = sum(like), num_votes = .N), by = 'id']

plot_histogram(x = log(ratings.like$num_likes + 1), breaks = 30) +
  scale_x_continuous(breaks = seq(0, 12, by = 1), limits = c(0, 12)) +
  scale_y_continuous(breaks = seq(0, 2e3, by = 2.5e2), limits = c(0, 2e3)) +
  labs(x = "Log Num. Likes", y = "Frequency") +
  theme_symplyr()
```

We got a very asymmetric empirical distribution, so we applied the `log` transformation to make it easy to visualize the full distribution. Just to have an idea, `10` in the log scale corresponds to about `22025` likes.

Now, let's see how the number of dislikes behave:

```{r}
ratings.like[, num_dislikes := num_votes - num_likes]

plot_histogram(x = log(ratings.like$num_dislikes + 1), breaks = 30) +
  labs(x = "Log Num. Dislikes", y = "Frequency") +
  scale_x_continuous(breaks = seq(0, 12, by = 1)) +
  scale_y_continuous(breaks = seq(0, 2e3, by = 2.5e2)) +
  theme_symplyr()
```

Is there any movie with more dislikes them likes?

```{r}
x <- log(ratings.like$num_likes + 1) - log(ratings.like$num_dislikes + 1)
plot_histogram(x = x, breaks = 30) +
  scale_x_continuous(breaks = seq(-4, 4)) +
  scale_y_continuous(breaks = seq(0, 1.5e3, by = 2.5e2)) +
  labs(x = "Log Num. Likes - Log Num. Dislikes", y = "Frequency") +
  theme_symplyr()
```

Since we have negative values, the answer to the last question is yes.

Is there a consistent bias in the average rating to popular movies?

```{r}
ggplot(ratings.like, aes(x = num_votes, y = avg_rating)) +
  geom_point(size = 0.5, alpha = 0.8, color = "gray20") +
  geom_smooth(color = "#66c2a5", alpha = 0.8, method = 'loess') +
  scale_x_continuous(breaks = seq(0, 1e5, by = 1e4)) +
  scale_y_continuous(breaks = seq(0, 5, by = 0.5)) +
  labs(x = "Num. Votes", y = "Avg. Rating") +
  theme_symplyr()
```

Again, the answer is yes. Movies with more than 20k votes have average rating above `3.0`. Movies with more than 50k votes tend to have average rating above `4.0`.
 
### Most Liked and Most Hated Movies

Which movies are the most liked by the users?

```{r}
movies <- file_import_csv("../data/movie_info.csv")

setorder(ratings.like, -num_likes)
best.movies <- merge(head(ratings.like, 10), movies[, c("id", "title")])
setorder(best.movies, -num_likes)

kbl(best.movies[, c('title', 'num_likes')], booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

How about the most disliked movies by the users?

```{r}
setorder(ratings.like, -num_dislikes)
worst.movies <- merge(head(ratings.like, 10), movies[, c("id", "title")])
setorder(worst.movies, -num_dislikes)

kbl(worst.movies[, c('title', 'num_dislikes')], booktabs = T) %>%
  kable_styling(latex_options = "striped")
```
### Genre Distribution

```{r}
genres <- file_import_csv("../data/movie_genre.csv")
genres <- unique(genres, by = c('id', 'genre_id'))

genre.prop <- genres[, .(n = .N), by = 'genre_name']

plot_bars(x = genre.prop$n, labels = genre.prop$genre_name) +
  labs(x = "Genre", y = "Num. Movies") +
  scale_y_continuous(breaks = seq(0, 1e4, by = 1e3)) +
  theme_symplyr()
```

### Most Liked Western and Horror Movies

Let's take a look at my personal favorite genres, Western and Horror. Which movies are the most liked by the audience?

**Western Movies:**

```{r}
genres.like <- merge(merge(ratings.like, genres, by = 'id'), movies, by = "id")
setorder(genres.like, -num_likes)

genres.like[genre_name == 'Western', c('title', 'num_likes')] %>% 
  head(10) %>% 
  kbl(booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

**Horror Movies:**

```{r}
genres.like[genre_name == 'Horror', c('title', 'num_likes')] %>% 
  head(10) %>% 
  kbl(booktabs = T) %>%
  kable_styling(latex_options = "striped")
```

I totally agree with the ranking!

### Most Liked and Most Hated Directors

```{r}
directors <- file_import_csv("../data/movie_producer.csv")[job == "Director",]
directors <- unique(directors, by = c("id", "person_id"))
directors <- merge(merge(directors, ratings.like, by = "id"), movies, by = "id")

directors.stats <- directors[, .(num_likes = sum(num_likes), num_dislikes = sum(num_dislikes), 
                                 best_movie = title[which.max(num_likes)], 
                                 worst_movie = title[which.max(num_dislikes)],
                                 name = person_name[1]), by = 'person_id']
directors.stats[, delta_like := num_likes - num_dislikes]
setorder(directors.stats, -delta_like)

directors.stats[, c('name', 'delta_like',  'best_movie')] %>% 
  head(15) %>% 
  kbl(booktabs = T, col.names = c('Director', 'Delta Likes', 'Most Liked Movie')) %>%
  kable_styling(latex_options = "striped")
```

```{r}
directors.stats[, c('name', 'delta_like',  'worst_movie')] %>% 
  tail(15) %>%
  dplyr::arrange(delta_like) %>% 
  kbl(booktabs = T, col.names = c('Director', 'Delta Likes', 'Most Hated Movie')) %>%
  kable_styling(latex_options = "striped")
```


### Most Liked and Most Hated Actors

```{r}
actors <- file_import_csv("../data/movie_actor.csv")
actors <- unique(actors, by = c("id", "actor_id"))
actors <- merge(merge(actors, ratings.like, by = "id"), movies, by = "id")

actors.stats <- actors[, .(num_likes = sum(num_likes), num_dislikes = sum(num_dislikes), 
                           best_movie = title[which.max(num_likes)], 
                           worst_movie = title[which.max(num_dislikes)],
                           name = actor_name[1]), by = 'actor_id']
actors.stats[, delta_like := num_likes - num_dislikes]
setorder(actors.stats, -delta_like)

actors.stats[, c('name', 'delta_like',  'best_movie')] %>% 
  head(15) %>% 
  kbl(booktabs = T, col.names = c('Actor', 'Delta Likes', 'Most Liked Movie')) %>%
  kable_styling(latex_options = "striped")
```
```{r}
actors.stats[, c('name', 'delta_like',  'worst_movie')] %>% 
  tail(15) %>% 
  dplyr::arrange(delta_like) %>% 
  kbl(booktabs = T, col.names = c('Actor', 'Delta Likes', 'Most Hated Movie')) %>%
  kable_styling(latex_options = "striped")
```

- Graphs from the notebook
- Sanity test: best and worst movies, actors and producers.

In this section, you will need to provide some form of visualization that summarizes or extracts a relevant characteristic or feature about the data. The visualization should adequately support the data being used. Discuss why this visualization was chosen and how it is relevant. Questions to ask yourself when writing this section:
- _Have you visualized a relevant characteristic or feature about the dataset or input data?_
- _Is the visualization thoroughly analyzed and discussed?_
- _If a plot is provided, are the axes, title, and datum clearly defined?_

## Algorithms and Techniques

- Algorithm description from the proposal

In this section, you will need to discuss the algorithms and techniques you intend to use for solving the problem. You should justify the use of each one based on the characteristics of the problem and the problem domain. Questions to ask yourself when writing this section:
- _Are the algorithms you will use, including any default variables/parameters in the project clearly defined?_
- _Are the techniques to be used thoroughly discussed and justified?_
- _Is it made clear how the input data or datasets will be handled by the algorithms and techniques chosen?_

## Benchmark

- Results for the user correlation

In this section, you will need to provide a clearly defined benchmark result or threshold for comparing across performances obtained by your solution. The reasoning behind the benchmark (in the case where it is not an established result) should be discussed. Questions to ask yourself when writing this section:
- _Has some result or value been provided that acts as a benchmark for measuring performance?_
- _Is it clear how this result or value was obtained (whether by data or by hypothesis)?_


# III. Methodology
_(approx. 3-5 pages)_

## Data Preprocessing

### Text Features

### Categorical Features

### Ratings

In this section, all of your preprocessing steps will need to be clearly documented, if any were necessary. From the previous section, any of the abnormalities or characteristics that you identified about the dataset will be addressed and corrected here. Questions to ask yourself when writing this section:
- _If the algorithms chosen require preprocessing steps like feature selection or feature transformations, have they been properly documented?_
- _Based on the **Data Exploration** section, if there were abnormalities or characteristics that needed to be addressed, have they been properly corrected?_
- _If no preprocessing is needed, has it been made clear why?_

## Implementation

- Show projections

### Content Based

### Collaborative Filtering

### Hybrid

In this section, the process for which metrics, algorithms, and techniques that you implemented for the given data will need to be clearly documented. It should be abundantly clear how the implementation was carried out, and discussion should be made regarding any complications that occurred during this process. Questions to ask yourself when writing this section:
- _Is it made clear how the algorithms and techniques were implemented with the given datasets or input data?_
- _Were there any complications with the original metrics or techniques that required changing prior to acquiring a solution?_
- _Was there any part of the coding process (e.g., writing complicated functions) that should be documented?_

## Refinement
In this section, you will need to discuss the process of improvement you made upon the algorithms and techniques you used in your implementation. For example, adjusting parameters for certain models to acquire improved solutions would fall under the refinement category. Your initial and final solutions should be reported, as well as any significant intermediate results as necessary. Questions to ask yourself when writing this section:
- _Has an initial solution been found and clearly reported?_
- _Is the process of improvement clearly documented, such as what techniques were used?_
- _Are intermediate and final solutions clearly reported as the process is improved?_


# IV. Results
_(approx. 2-3 pages)_

## Model Evaluation and Validation
In this section, the final model and any supporting qualities should be evaluated in detail. It should be clear how the final model was derived and why this model was chosen. In addition, some type of analysis should be used to validate the robustness of this model and its solution, such as manipulating the input data or environment to see how the model’s solution is affected (this is called sensitivity analysis). Questions to ask yourself when writing this section:
- _Is the final model reasonable and aligning with solution expectations? Are the final parameters of the model appropriate?_
- _Has the final model been tested with various inputs to evaluate whether the model generalizes well to unseen data?_
- _Is the model robust enough for the problem? Do small perturbations (changes) in training data or the input space greatly affect the results?_
- _Can results found from the model be trusted?_

### Ranking Correlation

### Precision and Recall

### Popularity Bias

## Justification
In this section, your model’s final solution and its results should be compared to the benchmark you established earlier in the project using some type of statistical analysis. You should also justify whether these results and the solution are significant enough to have solved the problem posed in the project. Questions to ask yourself when writing this section:
- _Are the final results found stronger than the benchmark result reported earlier?_
- _Have you thoroughly analyzed and discussed the final solution?_
- _Is the final solution significant enough to have solved the problem?_

### Correlation between Content Based and Collaborative Filtering Embedding

# V. Conclusion
_(approx. 1-2 pages)_

## Free-Form Visualization
In this section, you will need to provide some form of visualization that emphasizes an important quality about the project. It is much more free-form, but should reasonably support a significant result or characteristic about the problem that you want to discuss. Questions to ask yourself when writing this section:
- _Have you visualized a relevant or important quality about the problem, dataset, input data, or results?_
- _Is the visualization thoroughly analyzed and discussed?_
- _If a plot is provided, are the axes, title, and datum clearly defined?_

## Reflection
In this section, you will summarize the entire end-to-end problem solution and discuss one or two particular aspects of the project you found interesting or difficult. You are expected to reflect on the project as a whole to show that you have a firm understanding of the entire process employed in your work. Questions to ask yourself when writing this section:
- _Have you thoroughly summarized the entire process you used for this project?_
- _Were there any interesting aspects of the project?_
- _Were there any difficult aspects of the project?_
- _Does the final model and solution fit your expectations for the problem, and should it be used in a general setting to solve these types of problems?_

## Improvement
In this section, you will need to provide discussion as to how one aspect of the implementation you designed could be improved. As an example, consider ways your implementation can be made more general, and what would need to be modified. You do not need to make this improvement, but the potential solutions resulting from these changes are considered and compared/contrasted to your current solution. Questions to ask yourself when writing this section:
- _Are there further improvements that could be made on the algorithms or techniques you used in this project?_
- _Were there algorithms or techniques you researched that you did not know how to implement, but would consider using if you knew how?_
- _If you used your final solution as the new benchmark, do you think an even better solution exists?_

# References

[1] [Google's Introduction to Recommendation Systems](https://developers.google.com/machine-learning/recommendation).

[2] [Building a real-time embeddings similarity matching system](https://cloud.google.com/solutions/machine-learning/building-real-time-embeddings-similarity-matching-system).

[3] [The Amazon Recommendations Secret to Selling More Online](http://rejoiner.com/resources/amazon-recommendations-secret-selling-online/).

[4] Anderson, Chris. The long tail: Why the future of business is selling more for less. Hyperion, 2006.

[5] [Overview: Extracting and serving feature embeddings for machine learning](https://cloud.google.com/solutions/machine-learning/overview-extracting-and-serving-feature-embeddings-for-machine-learning).